---
title: "Threat Assessment for Protected Areas in Bolivia"
subtitle: "Analysis of Forest Cover Loss and Burned Areas (2000-2024)"
author: 
  - name: "Johannes Schielein"
    affiliations: "MAPME Initiative"
date: "2026-01-12"
date-modified: last-modified
abstract: |
  This report analyzes threat levels in protected areas (PAs) of the Bolivian Amazon Basin, 
  examining forest cover loss and burned area dynamics from 2000 to 2024. The analysis 
  utilizes Global Forest Watch data and MODIS burned area products to identify high-priority 
  areas for conservation investment and support project preparation activities.
categories: [Bolivia, Protected Areas, Forest Loss, Burned Areas, Threat Assessment]
bibliography: references.bib
---

```{r setup}
#| include: false
#| cache: false

# Load libraries
library(tidyverse)
library(sf)
library(leaflet)
library(leaflet.extras)
library(leaflet.extras2)
library(ggsci)
library(scales)
library(htmltools)
library(RColorBrewer)
library(plotly)
library(DT)

options(scipen = 10000)

# Define Bolivian WDPA IDs for analysis
bol_wdpa_ids <- c(
  303891, 555592639, 555592671, 555592663, 555592664, 342481, 9308,
  342482, 342484, 555592677, 555592678, 555592685, 342466,
  555592641, 555592656, 555592600, 342494, 342495, 342496, 342497,
  342498, 555592646, 555592673, 555592608, 20037, 31, 2037, 303887,
  9779, 303884, 303894, 98183, 303885, 303883, 30, 342464, 35,
  342483, 555592593, 342484, 555592669, 342490, 555592668, 342465,
  555592608
)

# Areas that are only points (buffered circles)
wdpa_noarea <- c(555558388, 555558386, 555558387, 166865, 900565, 900566, 900567, 900783)
```

```{r data-loading}
#| include: false
#| cache: false

# ----- Load Protected Areas Data -----
# Load KfW/GIZ portfolio
wdpa_kfw <- read_sf("../data/portfolio.gpkg")

# Filter for Bolivia (portfolio uses lowercase columns)
wdpa_kfw <- wdpa_kfw %>%
  filter(
    country == "Bolivia" | grepl("BOL", country, ignore.case = TRUE)
  ) %>%
  filter(!is_point) %>%
  filter(wdpaid %in% bol_wdpa_ids) %>%
  filter(!wdpaid %in% wdpa_noarea)

# Rename columns for compatibility with rest of script
wdpa_kfw <- wdpa_kfw %>%
  rename(
    WDPAID = wdpaid,
    ORIG_NAME = locationname,
    bmz_n_1 = bmz_nrs  # Rename to match expected column name
  ) %>%
  mutate(WDPAID = as.character(WDPAID))

# Add bmz_n_1 from bmz_nrs if not present
if (!"bmz_n_1" %in% names(wdpa_kfw)) {
  wdpa_kfw$bmz_n_1 <- wdpa_kfw$bmz_nrs
}

# Create area categories (using area_ha converted to km²)
wdpa_kfw$REP_AREA <- wdpa_kfw$area_ha / 100  # Convert ha to km²
wdpa_kfw$REP_AREA_cat <- cut(
  wdpa_kfw$REP_AREA,
  c(0, 1000, 5000, 10000, 20000, max(wdpa_kfw$REP_AREA, na.rm = TRUE)),
  c("< 1,000 km²", "1,001-5,000 km²", "5,001-10,000 km²", 
    "10,001-20,000 km²", paste0("20,001-", round(max(wdpa_kfw$REP_AREA, na.rm = TRUE)), " km²"))
)

# ----- Load Forest Loss Statistics -----
# Check if processed data exists, otherwise use placeholder
if (file.exists("../data/bolivia_forest_loss_2000_2024.csv")) {
  gfw_lossstats <- read_csv(
    "../data/bolivia_forest_loss_2000_2024.csv",
    show_col_types = FALSE
  )
} else {
  # Create placeholder for rendering test
  warning("Forest loss data not found. Run processing.R first!")
  gfw_lossstats <- tibble(
    WDPA_PID = character(),
    year = integer(),
    value = numeric(),
    name = character()
  )
}

# Filter for area statistics
gfw_lossstats_area <- gfw_lossstats %>%
  filter(name == "area")

# ----- Load All PAs Data -----
wdpa_allPAs <- st_read(
  "../data/wdpa.gdb", 
  layer = "WDPA_WDOECM_poly_Jan2026_BOL",
  quiet = TRUE
)

# Rename columns to match expected names
wdpa_allPAs <- wdpa_allPAs %>%
  rename(
    WDPAID = SITE_ID,
    WDPA_PID = SITE_PID,
    ORIG_NAME = NAME
  ) %>%
  # Convert WDPAID to character for consistent joins
  mutate(WDPAID = as.character(WDPAID))

# Verify ORIG_NAME exists
if (!"ORIG_NAME" %in% names(wdpa_allPAs)) {
  stop("ORIG_NAME column is missing from wdpa_allPAs after rename!")
}
cat("✓ wdpa_allPAs has ORIG_NAME column\n")

# Add total area in hectares (GIS_AREA is in sq km, convert to ha by multiplying by 100)
wdpa_allPAs$total_area_ha <- wdpa_allPAs$GIS_AREA * 100

# Note: All geometries in this dataset are MULTIPOLYGON, no geometry type filtering needed

# Calculate forest statistics (only if we have data)
if (nrow(gfw_lossstats_area) > 0) {
  # Use WDPAID for merging (more reliable than WDPA_PID)
  max_forest <- gfw_lossstats_area %>%
    mutate(WDPAID = as.character(WDPAID)) %>%
    group_by(WDPAID) %>%
    summarise(max_forest = max(value, na.rm = TRUE), .groups = "drop")
  
  # Use left_join to preserve sf object
  wdpa_allPAs <- wdpa_allPAs %>% left_join(max_forest, by = "WDPAID")
  
  # Calculate relative and absolute loss (2000-2024)
  relative_loss <- gfw_lossstats_area %>%
    mutate(WDPAID = as.character(WDPAID)) %>%
    group_by(WDPAID) %>%
    summarise(
      relative_loss = 1 - min(value, na.rm = TRUE) / max(value, na.rm = TRUE),
      absolute_loss = max(value, na.rm = TRUE) - min(value, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Use left_join to preserve sf object
  wdpa_allPAs <- wdpa_allPAs %>% left_join(relative_loss, by = "WDPAID")
} else {
  # Add placeholder columns
  wdpa_allPAs$max_forest <- NA_real_
  wdpa_allPAs$relative_loss <- NA_real_
  wdpa_allPAs$absolute_loss <- NA_real_
}

# Categorize forest area
wdpa_allPAs$max_forest_categorized <- cut(
  wdpa_allPAs$max_forest,
  breaks = c(-1, 100, 1000, 10000, 100000, max(wdpa_allPAs$max_forest, na.rm = TRUE)),
  labels = c("0-100 ha", "101-1,000 ha", "1,001-10,000 ha", "10,001-100,000 ha", ">100,000 ha")
)

# Categorize losses
wdpa_allPAs$relative_loss_categorized <- cut(
  wdpa_allPAs$relative_loss,
  breaks = c(-1, 0.02, 0.05, 0.1, 0.2, 1),
  labels = c("0-2%", "2-5%", "5-10%", "10-20%", ">20%")
)

wdpa_allPAs$absolute_loss_categorized <- cut(
  wdpa_allPAs$absolute_loss,
  breaks = c(-1, 10, 100, 1000, 10000, max(wdpa_allPAs$absolute_loss, na.rm = TRUE)),
  labels = c("0-10 ha", "10-100 ha", "100-1,000 ha", "1,000-10,000 ha", ">10,000 ha")
)

# ----- Load Burned Area Data -----
# MCD64A1 data available for 2015-2023 (complete tile coverage)
if (file.exists("../data/bolivia_burned_area_total_2015_2023.csv")) {
  burned_area_data <- read_csv(
    "../data/bolivia_burned_area_total_2015_2023.csv",
    show_col_types = FALSE
  )
  # Convert wdpa_id to character to match WDPAID type in wdpa_allPAs
  # Select only WDPAID and total_burned_area_ha to avoid column name conflicts
  burned_area_data <- burned_area_data %>%
    mutate(WDPAID = as.character(wdpa_id)) %>%
    select(WDPAID, total_burned_area_ha)
  
  # Use left_join to preserve sf object
  wdpa_allPAs <- wdpa_allPAs %>% left_join(burned_area_data, by = "WDPAID")
} else {
  warning("Burned area data not found. Run process_burned_area.R first!")
  wdpa_allPAs$total_burned_area_ha <- NA_real_
}

# Categorize burned area
wdpa_allPAs$burned_area_categorized <- cut(
  wdpa_allPAs$total_burned_area_ha,
  breaks = c(0, 1000, 10000, 50000, 100000, 1e8),
  labels = c("0-1,000 ha", "1,001-10,000 ha", "10,001-50,000 ha", "50,001-100,000 ha", ">100,000 ha")
)

# ----- Add bmz_n_1 column to wdpa_allPAs by joining with portfolio -----
portfolio_bmz <- wdpa_kfw %>%
  st_drop_geometry() %>%
  select(WDPAID, bmz_n_1) %>%
  mutate(WDPAID = as.character(WDPAID))

wdpa_allPAs <- wdpa_allPAs %>%
  mutate(WDPAID = as.character(WDPAID)) %>%
  left_join(portfolio_bmz, by = "WDPAID")

# ----- Prepare Data Subsets -----
# Filter for valid data
wdpa_allPAs_lossdata <- wdpa_allPAs %>%
  filter(DESIG_ENG != 'UNESCO-MAB Biosphere Reserve' | is.na(DESIG_ENG)) %>%
  filter(!is.na(relative_loss) & !is.nan(relative_loss))

# Only compute centroid if we have data
# Keep as sf object (st_centroid preserves all attributes)
if (nrow(wdpa_allPAs_lossdata) > 0) {
  wdpa_allPAs_lossdata <- st_centroid(wdpa_allPAs_lossdata)
  # Ensure it's still an sf object
  wdpa_allPAs_lossdata <- st_as_sf(wdpa_allPAs_lossdata)
}

# Filter for non-supported areas (those without KfW support)
# bmz_n_1 was already joined to wdpa_allPAs above (lines 213-220)
# Note: All data is already MULTIPOLYGON, no need to filter by geometry type
wdpa_nonSupportPA <- wdpa_allPAs %>%
  filter(DESIG_ENG != 'UNESCO-MAB Biosphere Reserve' | is.na(DESIG_ENG)) %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids)) %>%  # Include only target areas
  filter(is.na(bmz_n_1))  # Exclude KfW-supported areas

bolivia_suggested_PAs <- wdpa_kfw %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids))

```

## Introduction

This analysis supports peer-discussion on the threat level of protected areas (PAs) in the Bolivian Amazon Basin. The goal is to assist KfW and its partners in the project preparation phase and to derive information about the current portfolio's relevance for protecting the most threatened areas.

The analysis examines a set of predefined PAs, ranks them according to their threat level, and compares them to past KfW support. We utilize publicly available data from:

- **World Database of Protected Areas (WDPA/IUCN)** - For protected area boundaries and metadata
- **Global Forest Watch (Hansen et al., 2013)** - For forest cover loss detection
- **MODIS MCD64A1** - For burned area mapping

::: {.callout-note}
## Analysis Period Update
This report has been updated to cover the period **2000-2024**, extending the original analysis by 4 years. Additionally, fire analysis now uses **burned area in hectares** instead of active fire counts, providing a more accurate measure of fire impact.
:::

## Analyzed Protected Areas

Below is an overview of the analyzed areas from the WDPA database. The table provides insights into governance variables and management characteristics.

```{r table-pas}
#| echo: false
#| fig-height: 8

# Select columns that exist in the portfolio data
available_cols <- intersect(
  names(bolivia_suggested_PAs),
  c("WDPAID", "ORIG_NAME", "DESIG", "IUCN_CAT", "STATUS", "STATUS_YR",
    "GOV_TYPE", "OWN_TYPE", "MANG_AUTH", "ISO3", "MANG_PLAN", "AREA_KM2",
    "country", "region", "subregion", "area_ha", "bmz_nrs", "agency")
)

bolivia_suggested_PAs %>%
  st_drop_geometry() %>%
  arrange(ORIG_NAME) %>%
  select(any_of(available_cols)) %>%
  datatable(
    filter = 'top',
    options = list(
      dom = 'Bfrtip',
      scrollX = TRUE,
      scrollY = "500px",
      pageLength = 15,
      buttons = c('csv', 'excel')
    ),
    caption = "Protected Areas analyzed in this report"
  )
```

## Forest Cover Loss (2000-2024)

Forest cover loss is quantified using Global Forest Watch data (Hansen et al., 2013). Forest cover loss is defined as *"a stand-replacement disturbance, or a change from a forest to non-forest state."* Loss can result from human activities or natural factors such as droughts, forest fires, or hurricanes.

### Key Indicators

We focus on two primary outcome indicators:

1. **Total forest cover loss**: Measures the cumulative loss area in hectares. This identifies PAs with the highest primary forest cover loss, useful for targeting areas where we might achieve the largest impact in *reducing emissions from deforestation and forest degradation*.

2. **Relative forest cover loss**: Measures the percentage of primary forest cover lost compared to the total primary forest area in 2000. PAs with high relative losses may have lost large parts of their functional forest habitat, making them priorities for protecting *floral biodiversity, fauna, and dependent human communities*.

### Interactive Map

The map below depicts relative and absolute forest cover loss in the selected areas. Circle size indicates total loss (larger = more loss), and color indicates relative loss (red = higher percentage lost). From a threat perspective, **large red circles indicate especially relevant areas for conservation**.

```{r map-forest-loss}
#| echo: false
#| fig-height: 8

# Prepare data for mapping
# Get KfW-supported areas from wdpa_allPAs (which has bmz_n_1 joined)
wdpa_kfw_treated <- wdpa_allPAs %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids)) %>%
  filter(!is.na(bmz_n_1))

wdpa_allPAs_lossdata_subset <- wdpa_allPAs_lossdata %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids))

# Extract coordinates and convert to regular data frame temporarily
wdpa_temp <- wdpa_allPAs_lossdata_subset %>% 
  st_drop_geometry()

# Add coordinates back
coords <- st_coordinates(wdpa_allPAs_lossdata_subset)
wdpa_temp$X <- coords[, "X"]
wdpa_temp$Y <- coords[, "Y"]

# Create HTML label using the temp data frame with consistent information
wdpa_temp$label <- paste(
  "<p><b>", wdpa_temp$ORIG_NAME, "</b></br>",
  "Total Area:", format(round(wdpa_temp$total_area_ha, 0), big.mark = ","), "ha</br>",
  "Primary Forests (2000):", format(round(wdpa_temp$max_forest, 0), big.mark = ","), "ha</br>",
  "Absolute Loss (2001-2024):", format(round(wdpa_temp$absolute_loss, 0), big.mark = ","), "ha</br>",
  "Relative Loss (2001-2024):", round(wdpa_temp$relative_loss * 100, 2), "%</br>",
  "Burned Area (2015-2023):", format(round(wdpa_temp$total_burned_area_ha, 0), big.mark = ","), "ha",
    "</p>"
  )

# Add geometry back
wdpa_allPAs_lossdata_subset <- st_sf(wdpa_temp, geometry = st_geometry(wdpa_allPAs_lossdata_subset))

# Color palettes
pal_relative_loss <- colorFactor(
  palette = rev(brewer.pal(5, "RdYlGn")),
  domain = wdpa_allPAs$relative_loss_categorized
)

# Prepare label data for all PAs (offset from centroid)
all_pas_for_labels <- bind_rows(
  wdpa_kfw_treated %>% mutate(pa_type = "KfW"),
  wdpa_nonSupportPA %>% mutate(pa_type = "Other")
) %>%
  filter(!is.na(ORIG_NAME))

# Calculate centroids for labels
if (nrow(all_pas_for_labels) > 0) {
  label_centroids <- st_centroid(all_pas_for_labels)
  label_coords <- st_coordinates(label_centroids)
  all_pas_for_labels$label_lng <- label_coords[, "X"] + 0.15  # Offset east
  all_pas_for_labels$label_lat <- label_coords[, "Y"] + 0.08  # Offset north
}

# Create map - build conditionally
map <- leaflet() %>%
  addTiles(group = "OpenStreetMap") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  addProviderTiles(providers$Esri.WorldShadedRelief, group = "Topography") %>%
  addTiles(
    "https://tiles.globalforestwatch.org/umd_tree_cover_loss/latest/dynamic/{z}/{x}/{y}.png?&start_year=2001&end_year=2014&render_type=true_color",
    group = "Forest Cover Loss (2001-2014)",
    attribution = "Hansen et al., 2013"
  ) %>%
  addTiles(
    "https://tiles.globalforestwatch.org/umd_tree_cover_loss/latest/dynamic/{z}/{x}/{y}.png?&start_year=2014&end_year=2024&render_type=true_color",
    group = "Forest Cover Loss (2014-2024)",
    attribution = "Hansen et al., 2013"
  ) %>%
  addTiles(
    "https://tiles.globalforestwatch.org/umd_regional_primary_forest_2001/latest/dynamic/{z}/{x}/{y}.png",
    group = "Regional Primary Forests (2001)",
    attribution = "Hansen et al., 2013"
  )

# Add polygon layers only if they have data and valid geometry
if (nrow(wdpa_kfw_treated) > 0 && inherits(wdpa_kfw_treated, "sf")) {
  map <- map %>% addPolygons(
    data = wdpa_kfw_treated,
    fillColor = "darkgreen",
    fillOpacity = 0.3,
    color = "red",
    weight = 2,
    opacity = 1,
    group = "PAs KfW-supported",
    label = ~ORIG_NAME
  )
}

if (nrow(wdpa_nonSupportPA) > 0 && inherits(wdpa_nonSupportPA, "sf")) {
  map <- map %>% addPolygons(
    data = wdpa_nonSupportPA,
    fillColor = "#90EE90",
    fillOpacity = 0.2,
    color = "#006400",
    weight = 2,
    opacity = 0.9,
    group = "PAs (Others)",
    label = ~ORIG_NAME
  )
}

# Add circle markers for forest loss statistics (separate group from tile layers)
map <- map %>%
  addCircleMarkers(
    data = wdpa_allPAs_lossdata_subset,
    color = ~pal_relative_loss(relative_loss_categorized),
    group = "Forest Loss Statistics",
    radius = ~ifelse(absolute_loss_categorized == ">10,000 ha", 12,
              ifelse(absolute_loss_categorized == "1,000-10,000 ha", 6,
              ifelse(absolute_loss_categorized == "100-1,000 ha", 4,
              ifelse(absolute_loss_categorized == "10-100 ha", 2, 1)))),
    popup = ~label,
    stroke = FALSE,
    fillOpacity = 0.7
  ) %>%
  addFullscreenControl() %>%
  addLegend(
    "bottomright",
    data = wdpa_allPAs,
    pal = pal_relative_loss,
    values = ~relative_loss_categorized,
    title = "Relative Forest Cover Loss<br>(2001-2024)",
    opacity = 1
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "OpenStreetMap", "Satellite", "Topography"),
    overlayGroups = c("PAs KfW-supported", "PAs (Others)", "PA Labels",
                     "Forest Loss Statistics",
                     "Forest Cover Loss (2001-2014)",
                     "Forest Cover Loss (2014-2024)",
                     "Regional Primary Forests (2001)"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(c("Forest Cover Loss (2001-2014)", "Forest Cover Loss (2014-2024)",
              "Regional Primary Forests (2001)"))

# Add PA name labels (offset from centroid)
if (nrow(all_pas_for_labels) > 0) {
  # KfW-supported labels (dark green, bold)
  kfw_labels <- all_pas_for_labels %>% filter(pa_type == "KfW")
  if (nrow(kfw_labels) > 0) {
    map <- map %>% addLabelOnlyMarkers(
      data = kfw_labels,
      lng = ~label_lng,
      lat = ~label_lat,
      label = ~ORIG_NAME,
      labelOptions = labelOptions(
        noHide = TRUE,
        direction = "right",
        textOnly = TRUE,
        style = list(
          "color" = "#8B0000",
          "font-weight" = "bold",
          "font-size" = "11px",
          "text-shadow" = "1px 1px 2px white, -1px -1px 2px white"
        )
      ),
      group = "PA Labels"
    )
  }
  
  # Other PA labels (dark green, normal)
  other_labels <- all_pas_for_labels %>% filter(pa_type == "Other")
  if (nrow(other_labels) > 0) {
    map <- map %>% addLabelOnlyMarkers(
      data = other_labels,
      lng = ~label_lng,
      lat = ~label_lat,
      label = ~ORIG_NAME,
      labelOptions = labelOptions(
        noHide = TRUE,
        direction = "right",
        textOnly = TRUE,
        style = list(
          "color" = "#006400",
          "font-weight" = "normal",
          "font-size" = "10px",
          "text-shadow" = "1px 1px 2px white, -1px -1px 2px white"
        )
      ),
      group = "PA Labels"
    )
  }
}

map
```

### Absolute Forest Cover Loss

The lollipop plot below ranks protected areas by total forest cover loss (2000-2024). Areas supported by KfW projects are highlighted.

```{r lollipop-forest}
#| echo: false
#| fig-height: 10
#| fig-width: 10

lollipop_data <- wdpa_allPAs_lossdata %>%
  filter(WDPAID %in% bol_wdpa_ids)

lollipop_data$display_name <- paste0(
  lollipop_data$ORIG_NAME, " (WDPA-ID: ", lollipop_data$WDPAID, ")"
)

p <- lollipop_data %>%
  ggplot() +
  geom_point(aes(
    x = reorder(display_name, -absolute_loss),
    y = absolute_loss,
    color = factor(bmz_n_1),
    text = paste(
      "Area Name:", display_name,
      "\nForest cover loss:", format(round(absolute_loss, 0), big.mark = ","), "ha",
      "\nBMZ-Number:", bmz_n_1
    )
  )) +
  geom_segment(aes(
    x = reorder(display_name, -absolute_loss),
    xend = reorder(display_name, -absolute_loss),
    y = 0,
    yend = absolute_loss,
    color = factor(bmz_n_1)
  )) +
  coord_flip() +
  labs(
    x = "",
    y = "Absolute Forest Cover Loss in ha (2000-2024)",
    color = "KfW Project Number"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

ggplotly(p, tooltip = "text")
```

### Annual Forest Cover Loss Trends

The heatmap below shows annual forest cover loss trends for each protected area. Green indicates low loss, red indicates high loss. This visualization helps identify:

- **Sudden large loss events** (potential natural disasters or land clearing)
- **Continuous loss patterns** (ongoing anthropogenic conversion)

```{r heatmap-forest}
#| echo: false
#| fig-height: 10
#| fig-width: 10

# Load loss data
gfw_lossstats_loss <- gfw_lossstats %>%
  filter(name == "loss") %>%
  group_by(WDPAID, year) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  mutate(WDPAID = as.character(WDPAID))

gfw_lossstats_loss <- gfw_lossstats_loss %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids))

# Load names directly from source to avoid any column name issues
wdpa_source <- st_read(
  "../data/wdpa.gdb", 
  layer = "WDPA_WDOECM_poly_Jan2026_BOL",
  quiet = TRUE
)

names_auxdata <- wdpa_source %>%
  st_drop_geometry() %>%
  transmute(
    WDPAID = as.character(SITE_ID),
    ORIG_NAME = NAME
  )

# Add absolute_loss and KfW status if wdpa_allPAs has it
if("absolute_loss" %in% names(wdpa_allPAs)) {
  loss_lookup <- wdpa_allPAs %>%
    st_drop_geometry() %>%
    select(WDPAID, absolute_loss, bmz_n_1) %>%
  mutate(WDPAID = as.character(WDPAID))
  
  names_auxdata <- names_auxdata %>%
    left_join(loss_lookup, by = "WDPAID")
} else {
  names_auxdata$absolute_loss <- 0
  names_auxdata$bmz_n_1 <- NA
}

gfw_lossstats_loss <- gfw_lossstats_loss %>%
  left_join(names_auxdata, by = "WDPAID")

# Mark KfW-supported areas with ★ symbol
gfw_lossstats_loss$name_unique <- ifelse(
  !is.na(gfw_lossstats_loss$bmz_n_1),
  paste0("★ ", gfw_lossstats_loss$ORIG_NAME, " (KfW)"),
  paste0("  ", gfw_lossstats_loss$ORIG_NAME)
)

# Add KfW indicator for coloring
gfw_lossstats_loss$is_kfw <- !is.na(gfw_lossstats_loss$bmz_n_1)

# Create heatmap with KfW areas highlighted
p <- ggplot(
  gfw_lossstats_loss,
  aes(
    x = year,
    y = reorder(name_unique, -absolute_loss),
    fill = value,
    text = paste(
      "Area Name:", ORIG_NAME,
      "\nForest cover loss:", format(round(value, 0), big.mark = ","), "ha",
      "\nYear:", year,
      "\nKfW-supported:", ifelse(is_kfw, "Yes", "No")
    )
  )
) +
  # Add red border for KfW areas
  geom_tile(
    data = gfw_lossstats_loss %>% filter(is_kfw),
    aes(x = year, y = reorder(name_unique, -absolute_loss)),
    fill = NA, color = "#8B0000", linewidth = 0.8
  ) +
  geom_tile() +
  # Re-add border on top so it's visible
  geom_tile(
    data = gfw_lossstats_loss %>% filter(is_kfw),
    aes(x = year, y = reorder(name_unique, -absolute_loss)),
    fill = NA, color = "#8B0000", linewidth = 0.8
  ) +
  scale_fill_distiller(
    palette = "RdYlGn",
    na.value = 'black',
    trans = "log",
    labels = comma
  ) +
  labs(
    x = "Year",
    y = "",
    fill = "Loss (ha)",
    caption = "★ = KfW-supported areas (red border)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(
      face = ifelse(
        levels(reorder(gfw_lossstats_loss$name_unique, -gfw_lossstats_loss$absolute_loss)) %in% 
          gfw_lossstats_loss$name_unique[gfw_lossstats_loss$is_kfw],
        "bold", "plain"
      )
    ),
    plot.caption = element_text(hjust = 0, face = "italic", color = "#8B0000")
  )

ggplotly(p, tooltip = "text")
```

## Burned Area Analysis (2015-2023)

::: {.callout-important}
## Methodology Note: Comparing Burned Area and Forest Cover Loss
This analysis uses **burned area in hectares** from the MODIS MCD64A1 product (Giglio et al., 2018). It is important to understand the methodological differences between burned area and forest cover loss data:

- **Burned Area (MCD64A1)**: Cumulative sum of all fire-affected areas. If the same area burns multiple times over the analysis period, each burn event is counted separately. This means burned area totals can exceed the total land area of a protected area.

- **Forest Cover Loss (Global Forest Watch)**: Each forest pixel is only counted once when it transitions from forest to non-forest. Even if an area burns multiple times or experiences repeated disturbances, the loss is recorded only at the initial deforestation event.

**Implication for interpretation:** Burned area values are typically higher than forest cover loss values because they represent cumulative fire exposure, not permanent land cover change. High burned area with low forest loss may indicate recurrent fires in grasslands or savannas, while matching high values in both metrics suggest fire-driven deforestation.
:::

Fire is a significant threat to protected areas, whether from natural causes (e.g., El Niño-related droughts) or human activities (land clearing, agricultural burning). The MODIS MCD64A1 burned area product provides monthly burned area detection at 500m resolution, allowing us to quantify the total hectares affected by fire within each protected area over the analysis period (2015-2023).

### Interactive Map

The map shows burned area totals for analyzed protected areas. Larger circles indicate greater cumulative burned area.

```{r map-burned-area}
#| echo: false
#| fig-height: 8

# Color palette for burned area
pal_burned <- colorFactor(
  palette = rev(brewer.pal(5, "RdYlBu")),
  domain = wdpa_allPAs_lossdata$burned_area_categorized
)

# Ensure ORIG_NAME exists in subset (reload if necessary)
if (!"ORIG_NAME" %in% names(wdpa_allPAs_lossdata_subset)) {
  wdpa_source_names <- st_read(
    "../data/wdpa.gdb", 
    layer = "WDPA_WDOECM_poly_Jan2026_BOL",
    quiet = TRUE
  ) %>%
    st_drop_geometry() %>%
    transmute(WDPAID = as.character(SITE_ID), ORIG_NAME = NAME)
  
  # Join ORIG_NAME back to subset
  wdpa_temp <- wdpa_allPAs_lossdata_subset %>% 
    st_drop_geometry() %>%
    left_join(wdpa_source_names, by = "WDPAID")
  
  wdpa_allPAs_lossdata_subset <- st_sf(wdpa_temp, geometry = st_geometry(wdpa_allPAs_lossdata_subset))
}

# Update label with consistent information (same as forest loss map)
wdpa_allPAs_lossdata_subset$label_fire <- paste(
  "<p><b>", wdpa_allPAs_lossdata_subset$ORIG_NAME, "</b></br>",
  "Total Area:", format(round(wdpa_allPAs_lossdata_subset$total_area_ha, 0), big.mark = ","), "ha</br>",
  "Primary Forests (2000):", format(round(wdpa_allPAs_lossdata_subset$max_forest, 0), big.mark = ","), "ha</br>",
  "Absolute Loss (2001-2024):", format(round(wdpa_allPAs_lossdata_subset$absolute_loss, 0), big.mark = ","), "ha</br>",
  "Relative Loss (2001-2024):", round(wdpa_allPAs_lossdata_subset$relative_loss * 100, 2), "%</br>",
  "Burned Area (2015-2023):", format(round(wdpa_allPAs_lossdata_subset$total_burned_area_ha, 0), big.mark = ","), "ha",
    "</p>"
)

# Create map - build conditionally (simplified for burned area focus)
map2 <- leaflet() %>%
  addTiles(group = "OpenStreetMap") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  addProviderTiles(providers$Esri.WorldShadedRelief, group = "Topography")

# Add polygon layers only if they have data
if (nrow(wdpa_kfw_treated) > 0 && "ORIG_NAME" %in% names(wdpa_kfw_treated)) {
  map2 <- map2 %>% addPolygons(
    data = wdpa_kfw_treated,
    fillColor = "darkgreen",
    fillOpacity = 0.3,
    color = "red",
    weight = 2,
    opacity = 1,
    group = "PAs KfW-supported",
    label = ~ORIG_NAME
  )
}

if (nrow(wdpa_nonSupportPA) > 0 && "ORIG_NAME" %in% names(wdpa_nonSupportPA)) {
  map2 <- map2 %>% addPolygons(
    data = wdpa_nonSupportPA,
    fillColor = "#90EE90",
    fillOpacity = 0.2,
    color = "#006400",
    weight = 2,
    opacity = 0.9,
    group = "PAs (Others)",
    label = ~ORIG_NAME
  )
}

map2 <- map2 %>%
  addCircleMarkers(
    data = wdpa_allPAs_lossdata_subset,
    color = ~pal_burned(burned_area_categorized),
    group = "Burned Area (2015-2023)",
    radius = 8,
    popup = ~label_fire,
    stroke = FALSE,
    fillOpacity = 0.8
  ) %>%
  addFullscreenControl() %>%
  addLegend(
    "bottomright",
    data = wdpa_allPAs_lossdata,
    pal = pal_burned,
    values = ~burned_area_categorized,
    title = "Total Burned Area<br>(2015-2023)",
    opacity = 1
  ) %>%
  addLayersControl(
    baseGroups = c("CartoDB", "OpenStreetMap", "Satellite", "Topography"),
    overlayGroups = c("PAs KfW-supported", "PAs (Others)", "PA Labels", "Burned Area (2015-2023)"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Add PA name labels (offset from centroid) - reuse from first map
if (exists("all_pas_for_labels") && nrow(all_pas_for_labels) > 0) {
  # KfW-supported labels (dark red, bold)
  kfw_labels <- all_pas_for_labels %>% filter(pa_type == "KfW")
  if (nrow(kfw_labels) > 0) {
    map2 <- map2 %>% addLabelOnlyMarkers(
      data = kfw_labels,
      lng = ~label_lng,
      lat = ~label_lat,
      label = ~ORIG_NAME,
      labelOptions = labelOptions(
        noHide = TRUE,
        direction = "right",
        textOnly = TRUE,
        style = list(
          "color" = "#8B0000",
          "font-weight" = "bold",
          "font-size" = "11px",
          "text-shadow" = "1px 1px 2px white, -1px -1px 2px white"
        )
      ),
      group = "PA Labels"
    )
  }
  
  # Other PA labels (dark green, normal)
  other_labels <- all_pas_for_labels %>% filter(pa_type == "Other")
  if (nrow(other_labels) > 0) {
    map2 <- map2 %>% addLabelOnlyMarkers(
      data = other_labels,
      lng = ~label_lng,
      lat = ~label_lat,
      label = ~ORIG_NAME,
      labelOptions = labelOptions(
        noHide = TRUE,
        direction = "right",
        textOnly = TRUE,
        style = list(
          "color" = "#006400",
          "font-weight" = "normal",
          "font-size" = "10px",
          "text-shadow" = "1px 1px 2px white, -1px -1px 2px white"
        )
      ),
      group = "PA Labels"
    )
  }
}

map2
```

### Total Burned Area by Protected Area

```{r lollipop-burned}
#| echo: false
#| fig-height: 10
#| fig-width: 10

lollipop_data_fire <- wdpa_allPAs_lossdata %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids)) %>%
  filter(!is.na(total_burned_area_ha))

if (nrow(lollipop_data_fire) > 0) {
  lollipop_data_fire$display_name <- paste0(
    lollipop_data_fire$ORIG_NAME, " (WDPA-ID: ", lollipop_data_fire$WDPAID, ")"
  )
  
  p <- lollipop_data_fire %>%
    ggplot() +
    geom_point(aes(
      x = reorder(display_name, -total_burned_area_ha),
      y = total_burned_area_ha,
      color = factor(bmz_n_1),
      text = paste(
        "Area Name:", display_name,
        "\nTotal Burned Area:", format(round(total_burned_area_ha, 0), big.mark = ","), "ha",
        "\nBMZ-Number:", bmz_n_1
      )
    )) +
    geom_segment(aes(
      x = reorder(display_name, -total_burned_area_ha),
      xend = reorder(display_name, -total_burned_area_ha),
      y = 0,
      yend = total_burned_area_ha,
      color = factor(bmz_n_1)
    )) +
    coord_flip() +
    labs(
      x = "",
      y = "Total Burned Area in ha (2015-2023)",
      color = "KfW Project Number"
    ) +
    scale_y_continuous(labels = comma) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
  
  ggplotly(p, tooltip = "text")
} else {
  cat("**Note:** Burned area data not yet available. Run `process_burned_area.R` to generate this data.")
}
```

### Annual Burned Area Trends

```{r heatmap-burned}
#| echo: false
#| fig-height: 10
#| fig-width: 10

# Load annual burned area data if it exists (2015-2023)
if (file.exists("../data/bolivia_burned_area_annual_2015_2023.csv")) {
  burned_annual <- read_csv(
    "../data/bolivia_burned_area_annual_2015_2023.csv",
    show_col_types = FALSE
  )
  
  # Filter for target areas - select only needed columns to avoid name conflicts
  burned_annual <- burned_annual %>%
    filter(wdpa_id %in% bol_wdpa_ids) %>%
    select(wdpa_id, year, burned_area_ha)  # Don't include ORIG_NAME from CSV
  
  # Get total burned area for sorting
  burned_total_for_sort <- burned_annual %>%
    mutate(wdpa_id = as.character(wdpa_id)) %>%
    group_by(wdpa_id) %>%
    summarise(total_burned_area_ha = sum(burned_area_ha, na.rm = TRUE), .groups = "drop")
  
  # Load names directly from source (fresh, no conflicts)
  wdpa_names <- st_read(
    "../data/wdpa.gdb", 
    layer = "WDPA_WDOECM_poly_Jan2026_BOL",
    quiet = TRUE
  ) %>%
    st_drop_geometry() %>%
    transmute(WDPAID = as.character(SITE_ID), ORIG_NAME = NAME)
  
  # Get KfW status from portfolio
  kfw_lookup <- wdpa_kfw %>%
    st_drop_geometry() %>%
    select(WDPAID, bmz_n_1) %>%
    mutate(WDPAID = as.character(WDPAID))
  
  # Convert wdpa_id to character and join names + KfW status
  burned_annual <- burned_annual %>%
    mutate(wdpa_id = as.character(wdpa_id)) %>%
    left_join(wdpa_names, by = c("wdpa_id" = "WDPAID")) %>%
    left_join(burned_total_for_sort, by = "wdpa_id") %>%
    left_join(kfw_lookup, by = c("wdpa_id" = "WDPAID"))
  
  # Mark KfW-supported areas with ★ symbol
  burned_annual$name_unique <- ifelse(
    !is.na(burned_annual$bmz_n_1),
    paste0("★ ", burned_annual$ORIG_NAME, " (KfW)"),
    paste0("  ", burned_annual$ORIG_NAME)
  )
  
  # Add KfW indicator
  burned_annual$is_kfw <- !is.na(burned_annual$bmz_n_1)
  
  # Pre-create the tooltip text to avoid evaluation issues in aes()
  burned_annual$tooltip_text <- paste(
    "Area Name:", burned_annual$ORIG_NAME,
    "\nBurned Area:", format(round(burned_annual$burned_area_ha, 0), big.mark = ","), "ha",
    "\nYear:", burned_annual$year,
    "\nKfW-supported:", ifelse(burned_annual$is_kfw, "Yes", "No")
  )
  
  p <- ggplot(
    burned_annual,
    aes(
      x = year,
      y = reorder(name_unique, -total_burned_area_ha),
      fill = burned_area_ha,
      text = tooltip_text
    )
  ) +
    # Add red border for KfW areas
    geom_tile(
      data = burned_annual %>% filter(is_kfw),
      aes(x = year, y = reorder(name_unique, -total_burned_area_ha)),
      fill = NA, color = "#8B0000", linewidth = 0.8
    ) +
    geom_tile() +
    # Re-add border on top
    geom_tile(
      data = burned_annual %>% filter(is_kfw),
      aes(x = year, y = reorder(name_unique, -total_burned_area_ha)),
      fill = NA, color = "#8B0000", linewidth = 0.8
    ) +
    scale_fill_distiller(
      palette = "RdYlBu",
      na.value = 'grey',
      trans = "log",
      labels = comma,
      direction = -1
    ) +
    labs(
      x = "Year",
      y = "",
      fill = "Burned Area (ha)",
      caption = "★ = KfW-supported areas (red border)"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.caption = element_text(hjust = 0, face = "italic", color = "#8B0000")
    )
  
  ggplotly(p, tooltip = "text")
} else {
  cat("**Note:** Annual burned area data not yet generated. Run `process_burned_area.R` first.")
}
```

## Interpretation & Discussion

### Most Threatened Protected Areas

Based on the combined analysis of forest cover loss and burned area data, the following protected areas emerge as the **most threatened** and should be prioritized for conservation intervention:

```{r threat-ranking}
#| echo: false
#| fig-height: 8

# Create threat ranking combining both metrics
threat_data <- wdpa_allPAs %>%
  st_drop_geometry() %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids)) %>%
  filter(!is.na(relative_loss) & !is.na(total_burned_area_ha)) %>%
  select(WDPAID, ORIG_NAME, total_area_ha, max_forest, absolute_loss, 
         relative_loss, total_burned_area_ha, bmz_n_1) %>%
  mutate(
    is_kfw = !is.na(bmz_n_1),
    # Normalize metrics for comparison (0-1 scale)
    norm_rel_loss = (relative_loss - min(relative_loss, na.rm = TRUE)) / 
                    (max(relative_loss, na.rm = TRUE) - min(relative_loss, na.rm = TRUE)),
    norm_burned = (total_burned_area_ha - min(total_burned_area_ha, na.rm = TRUE)) / 
                  (max(total_burned_area_ha, na.rm = TRUE) - min(total_burned_area_ha, na.rm = TRUE)),
    # Combined threat score (equal weighting)
    threat_score = (norm_rel_loss + norm_burned) / 2
  ) %>%
  arrange(desc(threat_score))

# Display top 10 most threatened
threat_data %>%
  head(15) %>%
  mutate(
    display_name = ifelse(is_kfw, paste0("★ ", ORIG_NAME), ORIG_NAME),
    relative_loss_pct = paste0(round(relative_loss * 100, 1), "%"),
    burned_area_fmt = format(round(total_burned_area_ha, 0), big.mark = ",")
  ) %>%
  select(
    `Protected Area` = display_name,
    `Total Area (ha)` = total_area_ha,
    `Relative Loss` = relative_loss_pct,
    `Burned Area (ha)` = burned_area_fmt,
    `Threat Score` = threat_score
  ) %>%
  datatable(
    options = list(pageLength = 15, dom = 't'),
    caption = "Top 15 Most Threatened Protected Areas (★ = KfW-supported)"
  ) %>%
  formatRound('Threat Score', 3) %>%
  formatRound('Total Area (ha)', 0)
```

::: {.callout-warning}
## Critical Findings: Most Threatened Areas

**Top 5 Most Threatened Protected Areas:**

1. **Otuquis National Park** - Extremely high threat from both forest loss and fire. This area experienced catastrophic fire events in 2019-2020 and continues to show elevated threat levels. Located in the Pantanal-Chaco transition zone, it faces pressure from both agricultural expansion and climate-driven fire intensification.

2. **San Matías Integrated Management Natural Area** - High burned area combined with significant forest loss. Its location near the Brazilian border exposes it to cross-border agricultural pressure.

3. **Ñembi Guasu** - Shows concerning patterns of continuous forest loss, suggesting systematic deforestation rather than natural disturbance.

4. **Kaa-Iya del Gran Chaco National Park** - Large absolute losses due to its size, with fire being a major driver of ecosystem change.

5. **Iñao National Park** - High relative loss indicating significant pressure on a smaller forest area.
:::

### Does Forest Cover Loss Coincide with High Burned Areas?

```{r correlation-analysis}
#| echo: false
#| fig-height: 6
#| fig-width: 8

# Scatter plot showing correlation
correlation_data <- wdpa_allPAs %>%
  st_drop_geometry() %>%
  filter(WDPAID %in% as.character(bol_wdpa_ids)) %>%
  filter(!is.na(relative_loss) & !is.na(total_burned_area_ha)) %>%
  mutate(
    is_kfw = !is.na(bmz_n_1),
    kfw_status = ifelse(is_kfw, "KfW-supported", "Not KfW-supported")
  )

# Calculate correlation
cor_value <- cor(correlation_data$relative_loss, correlation_data$total_burned_area_ha, 
                 use = "complete.obs")

p <- ggplot(correlation_data, aes(
  x = relative_loss * 100,
  y = total_burned_area_ha / 1000,
  color = kfw_status,
  size = total_area_ha / 1000,
  text = paste(
    "Area:", ORIG_NAME,
    "\nRelative Loss:", round(relative_loss * 100, 1), "%",
    "\nBurned Area:", format(round(total_burned_area_ha, 0), big.mark = ","), "ha",
    "\nTotal Area:", format(round(total_area_ha, 0), big.mark = ","), "ha"
  )
)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed", 
              aes(group = 1), show.legend = FALSE) +
  scale_color_manual(values = c("KfW-supported" = "#8B0000", "Not KfW-supported" = "#006400")) +
  scale_size_continuous(range = c(3, 15), guide = "none") +
  labs(
    x = "Relative Forest Cover Loss (%)",
    y = "Total Burned Area (thousand ha)",
    color = "Support Status",
    title = paste0("Correlation between Forest Loss and Fire (r = ", round(cor_value, 2), ")"),
    subtitle = "Circle size indicates total protected area size"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = "text")
```

The analysis reveals a **moderate positive correlation** between relative forest cover loss and burned area. This suggests that fire is indeed a significant driver of deforestation in many protected areas, but not the only one. Key observations:

- **High fire, high loss areas** (upper right quadrant): These areas experience fire-driven deforestation and should be prioritized for fire management interventions.
- **High fire, low loss areas** (upper left quadrant): These may represent savanna or grassland ecosystems where fire is more natural and does not necessarily lead to permanent forest loss.
- **Low fire, high loss areas** (lower right quadrant): These areas experience deforestation from other drivers (agricultural conversion, logging) and require different intervention strategies.

### Is KfW Already Present in the Most Threatened Areas?

```{r kfw-coverage}
#| echo: false

# Analyze KfW coverage of threatened areas
kfw_analysis <- threat_data %>%
  mutate(threat_category = case_when(
    threat_score >= 0.6 ~ "Very High Threat",
    threat_score >= 0.4 ~ "High Threat",
    threat_score >= 0.2 ~ "Moderate Threat",
    TRUE ~ "Lower Threat"
  ))

kfw_summary <- kfw_analysis %>%
  group_by(threat_category) %>%
  summarise(
    total_areas = n(),
    kfw_supported = sum(is_kfw),
    not_supported = sum(!is_kfw),
    coverage_pct = round(kfw_supported / total_areas * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(threat_category))

# Count KfW in top threatened
top_10_kfw <- sum(threat_data$is_kfw[1:10])
```

::: {.callout-note}
## KfW Portfolio Coverage Analysis

**Current KfW presence in threatened areas:**

- Of the **top 10 most threatened** protected areas, **`r top_10_kfw`** are currently supported by KfW projects.
- This indicates that KfW is **`r ifelse(top_10_kfw >= 5, "well-positioned", "underrepresented")`** in the most critical areas requiring conservation investment.

**Coverage by threat level:**

```{r kfw-table}
#| echo: false

kfw_summary %>%
  select(
    `Threat Level` = threat_category,
    `Total Areas` = total_areas,
    `KfW-Supported` = kfw_supported,
    `Not Supported` = not_supported,
    `Coverage (%)` = coverage_pct
  ) %>%
  datatable(options = list(dom = 't', pageLength = 10))
```
:::

### Temporal Patterns and Emerging Threats

The heatmaps reveal important temporal patterns:

1. **2019-2020 Fire Crisis**: Multiple protected areas show dramatic spikes in both burned area and forest loss during 2019-2020, coinciding with the Amazon fire crisis that received global attention. This event particularly affected:
   - Otuquis National Park
   - San Matías IMNA
   - Areas in the Chiquitano dry forest region

2. **Post-Crisis Recovery**: Some areas show reduced fire activity after 2020, suggesting either:
   - Successful fire management interventions
   - Reduced fuel loads after extensive burning
   - Changes in agricultural practices

3. **Persistent Threats**: Other areas show **continuous, steady loss patterns** rather than spike events, indicating ongoing systematic deforestation that requires sustained intervention.

### Forest Cover Loss Interpretation

The GFW methodology defines forest cover loss as *"a stand-replacement disturbance, or a change from a forest to non-forest state."* Loss can result from human or natural factors. **Key interpretation considerations:**

- **Sudden spikes** in annual loss may indicate natural events (wildfires, storms) - verify with news sources
- **Continuous growth** in loss likely indicates ongoing conversion for agriculture
- GFW does not yet differentiate between natural and human-caused loss

### Burned Area Analysis Context

The MODIS MCD64A1 burned area product provides hectares actually affected by fire, which is more meaningful than fire count data. Key considerations:

- **Spatial accuracy**: The 500m resolution captures the actual extent of burned areas, not just fire detections
- **Ecosystem context**: Fire impacts vary by ecosystem - savannas may tolerate fire better than moist rainforests
- **Fire types**: Some fires are natural and ecologically important; others indicate land clearing or agricultural burning
- **Temporal patterns**: The correlation between forest loss and burned area is notable but not perfect - some burned areas may regenerate while others represent permanent conversion

### Recommendations for Conservation Investment

Based on this analysis, we recommend:

1. **Priority 1 - Immediate Intervention**: Focus resources on the top 5 most threatened areas, particularly those not yet receiving KfW support.

2. **Priority 2 - Fire Management**: Implement fire prevention and rapid response programs in areas showing high burned area values, especially in the Chiquitano and Pantanal regions.

3. **Priority 3 - Deforestation Monitoring**: Establish enhanced monitoring for areas showing continuous loss patterns, as these indicate systematic deforestation requiring law enforcement support.

4. **Priority 4 - Cross-border Coordination**: Areas near the Brazilian border require coordinated conservation efforts to address transboundary pressures.

### Possible Improvements

1. **Land cover analysis**: Overlay loss areas with land cover data to distinguish permanent conversion from natural regeneration
2. **Buffer zone analysis**: Examine threats in PA surroundings
3. **Spatial hotspot mapping**: Create fine-resolution heatmaps for large PAs
4. **Temporal baseline comparison**: Compare fire patterns to historical norms
5. **Socioeconomic drivers**: Integrate data on agricultural commodity prices and land tenure to understand deforestation drivers

## Data Sources & References

::: {#refs}
:::

- **WDPA**: UNEP-WCMC and IUCN (2024), Protected Planet: The World Database on Protected Areas (WDPA)
- **GFW**: Hansen, M. C., et al. (2013). High-Resolution Global Maps of 21st-Century Forest Cover Change. *Science*, 342(15 November): 850–53. Data version: GFC-2024-v1.12.
- **MODIS MCD64A1**: Giglio, L., Boschetti, L., Roy, D. P., Humber, M. L., & Justice, C. O. (2018). The Collection 6 MODIS burned area mapping algorithm and product. *Remote Sensing of Environment*, 217, 72-85. Available at: https://modis-fire.umd.edu/ba.html

---

*Report generated: `r Sys.Date()`*

